\chapter{GIT usage}

This report is usable both from the web-interface overleaf \url{https://www.overleaf.com/project/5ebea6ecfa55c300015e6073} and as git repository from: \url{https://git.overleaf.com/5ebea6ecfa55c300015e6073}

If you have problems accessing/writing to github or overleaf, contact \href{mailto:Heiko.Klein@met.no}{Heiko.Klein@met.no}

\section{Report usage from overleaf}

Login to overleaf. Make the change you want. Press Ctrl-s or 'Recompile' regularly so you don't introduce too many bugs. Your changes are regularly saved automatically.

\section{Report usage via git+overleaf}

Make sure you have a password set on overleaf. If you don't know your password (e.g. because you login with google), reset the password. You should cache your credentials to avoid typing them all the time (this will cache your username/password for git in memory):
\begin{verbatim}
git config credential.helper 'cache'
\end{verbatim}


Download the report once:
\begin{verbatim}
    git clone https://git.overleaf.com/5ebea6ecfa55c300015e6073 Report\_2020
    # enter username/password
\end{verbatim}
All further work will be done in the directory Report\_2020. Remember that with git, you are working with at least 3 versions, a working-copy (where you make your changes), a local repository (add/status/commit) and the original repository on overleaf (pull/push).

To get the latest version of the report, run from within above directory:
\begin{verbatim}
    git pull
\end{verbatim}

To create a local version of your changes add and commit the to your local repository:

\begin{verbatim}
    git add file1.tex file2.tex
    git commit -m 'changed a bit'
\end{verbatim}

Check regularly if you haven't forgotten to add any files to your local repository:
\begin{verbatim}
    git status
\end{verbatim}

Make your local repository available to everybody else. (Ensure you are working on the latest version. And if conflicts, you should fix them and prepare a new version). Make this when nobody else is working on the report, otherwise, there will be changes all the time between pull and push.
\begin{verbatim}
    git pull # accept merging changes
    git push
\end{verbatim}

More information on using github and overleaf can be found here: \href{https://confluence.desy.de/download/attachments/97009435/overleaf_git_howto.pdf?version=1&modificationDate=1550237113964&api=v2}.
